### 2020-09-09 Manuscript/Internal_Review.tex

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal to be T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%). {\color{blue} I couldn't find where you were doing this calculation, so I made one up -- see lines 1134 and following in the Rmd. BUT ALSO : community 3 isn't sensitive to pH, so what are we even saying here?}

\\

%Under a common environmental condition, for example, in Hood Canal in summer, we expect such a planktonic community to go from occurring less than 1\% of the time in 2017 (0.001\% at T $=$ 15$\degree$C and pH $=$ 8.4; conditions more extreme than this occur ca. 50\% of the time) to occurring 37.5\% of the time under an equivalently-common condition in 2095 (84\% at T $=$ 22$\degree$C and pH $=$ 8.05).

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, averaged over all environmental conditions, Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.5 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (p $=$ 0.8).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1179 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Top_fig3_2020-09-03.Other.sel.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea {\color{red} for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panels shows:  (\textbf{.1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{.2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{.3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? I went through line by line; double check calculations -- see lines 1207 and following in the Rmd. Plus, a reviewer wanted a citation of the Coscinodiscus statement, so I added some.}

{\color{blue}The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).}

@@

{\color{blue} needs RAMON review}
Among the taxa surveyed, diatoms are of particular interest for their ubiquity in the world's oceans and their important roles in marine food webs \cite{armbrust2009life, wasmund2017diatom}, as well as in ecological and evolutionary theory \cite{margalef11978}. Our model suggests that diatoms will decrease in richness between the present and 2095, particularly in the Hood Canal, where extreme temperatures are more common. Although the most prevalent response among diatoms is a decrease in suitability, some substantial variability in responses exists within the group. For example, the centric diatom \textit{Coscinodiscus} spp., which is a food source for \textit{Acartia} spp. copepods \cite{jansen2008copepods} and many other animal species, will see future suitable habitat only in colder waters such as those in San Juan Island, while \textit{Skeletonema} spp. and the harmful algal bloom (HAB)-forming species \textit{Pseudo-nitzschia} spp. will see their habitat suitability remain constant or slightly increased, especially at low pH levels. {\color{blue} is this in supplement?}

>>>>

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to present the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).

% Under a common environmental condition, for example, in Hood Canal in summer, we expect such a planktonic community to go from occurring less than 1\% of the time in 2017 (0.001\% at T $=$ 15$\degree$C and pH $=$ 8.4; conditions more extreme than this occur ca. 50\% of the time) to occurring 37.5\% of the time under an equivalently-common condition in 2095 (84\% at T $=$ 22$\degree$C and pH $=$ 8.05).

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, 95\% confidence interval in median species richness  -0.08, +0.04; Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 1.84-2.2, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.47-0.57 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (median change 0-0.08 p $=$ 0.04).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$).

%%%%%
 \begin{figure*}
 \centering
 \includegraphics[width=1\linewidth]{Figures_for_MS/Top_fig3_2020-09-03.Other.sel.png}
\caption{\footnotesize{Forecasted changes in plankton in the Salish Sea for Dinoflagellates (panel \textbf{A}), Diatoms (\textbf{B}) and Haptophytes (\textbf{C}). Each panels shows:  (\textbf{.1}) probability densities for the occurrence of selected taxa (species- and genus-level) for 2017 (blue) and 2095 (red); data are mean probabilities over 100 model draws, and variance in probability is due to differences in underlying environmental conditions.  (\textbf{.2}) Changes in relative species richness within each phylum across the simulated scenario. (\textbf{.3}) Relative taxon richness (raster color, warmer colors are more taxon-rich) for each of these same higher-taxa, for plausible ranges of pH and Temperature. Envelopes of observed (2017, blue) and modeled (2095, red) annual conditions in the Salish Sea shown for reference. Hood Canal and San Juan Island plotted separately to illustrate environmental differences between them.}}

@@

The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).

@@

Among the taxa surveyed, diatoms are of particular interest for their ubiquity in the world's oceans and their important roles in marine food webs \cite{armbrust2009life, wasmund2017diatom}, as well as in ecological and evolutionary theory \cite{margalef11978}. Our model suggests that diatoms will decrease in richness between the present and 2095, particularly in the Hood Canal, where extreme temperatures are more common. Although the most prevalent response among diatoms is a decrease in suitability, some substantial variability in responses exists within the group. For example, the centric diatom \textit{Coscinodiscus} spp., which is a food source for \textit{Acartia} spp. copepods \cite{jansen2008copepods} and many other animal species, will see future suitable habitat only in colder waters such as those in San Juan Island, while \textit{Skeletonema} spp. and the harmful algal bloom (HAB)-forming species \textit{Pseudo-nitzschia} spp. will see their habitat suitability remain constant or slightly increased, especially at low pH levels (see Supplemental Material).

### 2020-09-09 Manuscript/What.lies.ahead.reviewed.tex

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal to be T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%). {\color{blue} I couldn't find where you were doing this calculation, so I made one up -- see lines 1134 and following in the Rmd. BUT ALSO : community 3 isn't sensitive to pH, so what are we even saying here?}

\\

%Under a common environmental condition, for example, in Hood Canal in summer, we expect such a planktonic community to go from occurring less than 1\% of the time in 2017 (0.001\% at T $=$ 15$\degree$C and pH $=$ 8.4; conditions more extreme than this occur ca. 50\% of the time) to occurring 37.5\% of the time under an equivalently-common condition in 2095 (84\% at T $=$ 22$\degree$C and pH $=$ 8.05).

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, averaged over all environmental conditions, Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.5 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (p $=$ 0.8).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1179 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? I went through line by line; double check calculations -- see lines 1207 and following in the Rmd. Plus, a reviewer wanted a citation of the Coscinodiscus statement, so I added some.}

{\color{blue}The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).}

>>>>

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal {\color{red} in 2095 to present the conditions in which community 3 is the most likely community 75\% of the time, an increase of 22\% compared to 2017(see Supplemental Material).} \st{T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%).} {\color{blue} I couldn't find where you were doing this calculation, so I made one up -- see lines 1134 and following in the Rmd. BUT ALSO : community 3 isn't sensitive to pH, so what are we even saying here?}

% Under a common environmental condition, for example, in Hood Canal in summer, we expect such a planktonic community to go from occurring less than 1\% of the time in 2017 (0.001\% at T $=$ 15$\degree$C and pH $=$ 8.4; conditions more extreme than this occur ca. 50\% of the time) to occurring 37.5\% of the time under an equivalently-common condition in 2095 (84\% at T $=$ 22$\degree$C and pH $=$ 8.05).

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, 95\% confidence interval in median species richness  -0.08, +0.04; Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 1.84-2.2, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.47-0.57 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (median change 0-0.08 p $=$ 0.04).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1179 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? I went through line by line; double check calculations -- see lines 1207 and following in the Rmd. Plus, a reviewer wanted a citation of the Coscinodiscus statement, so I added some. RYAN- I agree with the statement. Now we do these calculations in the script. I will add that the last bits of the prevalence of Alexandrium and Coscinoduscus are impactful and supercool}

{\color{red}The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).}
